---
title: "Baseball Level Analysis"
author: "Jaden Thomas"
date: "2024-02-05"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Problem Description
To be able to determine what level of baseball a player plays at by the individuals height and weight, a logistic regression model is fit.

The model formula is:

$$p(X)=\frac{e^{\beta_0+\beta_1HT_i+\beta_2WT_i}}{1+e^{\beta_0+HT_i+WT_i}}; i=1,...,n$$
Or,
$$\log(\frac{p(X)}{1-p(X)})=\beta_0+\beta_1HT_i+\beta_2WT_i; i=1,...n$$

Where HT is in inches, WT is in lbs, and the two classes are MLB and Club, with MLB labeled as 1 and Club baseball labeled as 0.

# Import Necessary Packages
```{r, warning=F, message=F}
library(tidyverse)
library(ggthemes)
library(verification)
```

# Set Seed
```{r}
set.seed(2024)
```

# Import Datasets
```{r, message=F}
mlb_2023 <- read_csv("mlbBaseballPlayers_2023.csv")
club_2024 <- read_csv("clubBaseballPlayers_2024.csv")

```

# Format Data
```{r}
club_2024 <- club_2024 %>% 
  mutate(Age=(year(Sys.Date())-year(as.Date(paste0("01/", club_2024$DOB), format="%d/%m/%Y"))),
         POS=str_split_i(POS, pattern=" / ", i=1)) %>% 
  rename(BAT=Bats, THW=Throws)
full <- bind_rows(mlb_2023, club_2024)
sub <- full %>% dplyr::select(fname, lname, Age, HT, WT, city, state, POS, Team, League) %>% 
  mutate(League=as.factor(League)) %>% mutate(MLB=(League=="MLB"))
```

# EDA
```{r}
ggplot(sub, aes(x=WT, y=HT)) +
  geom_point(aes(color=League)) +
  labs(title="Height and Weight of Players by League", x="Weight in lbs",
       y="Height in inches") +
  theme_economist()

sub %>% ggplot(aes(x=Age, y=League)) +
  geom_boxplot(fill="#e63946") +
  theme_economist() +
  labs(title="Player Age by League", x="Age")

sub %>% ggplot(aes(x=HT, y=League)) +
  geom_boxplot(fill="#e63946") +
  theme_economist() +
  labs(title="Player Height by League", x="Height")

sub %>% ggplot(aes(x=WT, y=League)) +
  geom_boxplot(fill="#e63946") +
  theme_economist() +
  labs(title="Player Weight by League", x="Weight")
```

```{r}
summary(sub)
sub %>% group_by(League) %>% summarise(meanHT=mean(HT), meanWT=mean(WT), meanAGE=mean(Age), n=n())
```

As seen by the graphs and numerical summary, the average height and weight for MLB players seems to be higher than that of Club baseball players.


# Logistic Regression Model

## Train Test Split

```{r}
train <- sample(c(TRUE, FALSE), nrow(sub), replace=T, prob=c(0.7, 0.3))
sub.train <- sub[train,]
sub.test <- sub[!train,]
Y.test <- sub.test$MLB
```


## Model Fitting

```{r}
m.fit <- glm(MLB~HT+WT, data=sub.train, family="binomial")
```


## Model Summary

```{r}
summary(m.fit)
plot(m.fit)
```

The fitted model is $\hat{p_i}=logit(-19.109369+0.136833*HT_i+0.047340*WT_i); i=1,...,n$. Both predictors HT and WT are significant also with the null hypothesis $H_0:\beta_i=0$ being rejected for all i=(1,2,3).

## Model Prediction

```{r}
# T means that, yes the individual is in the MLB
m.probs <- predict(m.fit, sub.test, type="response") 
m.pred <- rep(F, length(m.probs))
m.pred[m.probs>.5] <- T
```


## Model Evaluation

```{r}
table(m.pred, Y.test, dnn=c("Predicted MLB", "Actual MLB"))
mean(m.pred==Y.test)

roc.plot(x=as.numeric(Y.test), pred=as.numeric(m.pred), main="ROC curve for MLB Classifcation", plot.thres = NULL)
```


The model performs with a 79.43925% accuracy, a sensitivity of 77.211796% and a specificity of 81.648936.

## Conclusion

As the prediction results from the logistic regression show, predicting the level at which an individual play baseball at, either MLB or Club, can be done at a relatively high rate with just the height and weight of the players as independent variables using logistic regression.
